### 2015-09-04 ms_radial_092713/ms_radial_chapter.tex
chapter 2 summary
fixed figure placement
spell checked

\chapter{Fast Radial Flows in Transition Disk Holes}\label{c:radial_flows}

\section*{abstract}

@@

A subset of T Tauri and Herbig Ae disks have large, mass-depleted cavities in 
their inner regions, inferred from their infrared spectra 
\citep[e.g.,][]{Kim13} or spatially resolved imaging at infrared 
\citep[e.g.,][]{geers2007} and mm wavelengths \citep[e.g.,][]{brown09}.  The
cavities in these ``transition'' disks have radii ranging from $\sim$4\,AU 
\citep[TW Hya;][]{calvet02,hughes07} to 20\,AU \citep[GM Aur;][]{hughes09} to 
40--50\,AU \citep[LkCa 15;][]{andrews11b,isella12} to 140\,AU \citep[HD 
142527;][]{casassus13}.  The suspicion is that dynamical interactions with 
nascent planetary systems evacuate the inner disk regions 
\citep[e.g.,][]{Zhu11}.

@@

Disk radial velocities driven by embedded massive planets or brown dwarfs can 
be healthy fractions of orbital (azimuthal) velocities.  We derive a crude 
estimate\footnote{Our ``zero-dimensional'' analysis ignores the fact that 
actual disk flows around planets are two-dimensional and dominated by
non-axisymmetric streams flowing in and around planetary Hill spheres.
Still, some indirect empirical support for the scalings in equation 
(\ref{eqn:simple}) has been found by \cite{fung13}.
}
 for $v_R$ by equating the one-sided Lindblad torque --- the amount of 
angular momentum transmitted outward across a planet's orbit 
\citep{goldreich80} --- with the angular momentum advected inward by accretion, 
$\dot{M} v_{\rm K} R$.  This balance yields
\begin{equation}
v_R \sim \left( \frac{M_p}{M_\ast} \right)^2 \left( \frac{H}{R} \right)^3 v_{\rm K} \label{eqn:simple}
\end{equation}
where we have used the total linear Lindblad torque, integrated over all 
resonances outside the torque cut-off zone at distances $\gtrsim H$ away from 
the planet (see, e.g., \citealt{goodman01} or \citealt{crida06}, and references 
therein).  For a planet mass $M_p \sim 10\,M_{\rm Jup}$, a host stellar mass 
$M_\ast \sim 1\,M_\odot$, and a typical disk aspect ratio $H/R \sim 0.1$, 
equation (\ref{eqn:simple}) evaluates to
\begin{equation}
v_R \sim 0.1 \,v_{\rm K} \sim 0.5 \left( \frac{30\,{\rm AU}}{R} \right)^{1/2}\,{\rm km} \,\, {\rm s}^{-1}. 
\label{eqn:simple1}
\end{equation}
Modern heterodyne receivers on (sub)millimeter wavelength interferometers 
routinely observe disks at $\sim$0.1 km s$^{-1}$ resolution, and the Atacama 
Large Millimeter Array (ALMA) is expected to provide high sensitivity data at 
finer velocity scales yet.  Evidence for free-fall radial velocities can be 
found in the HCO$^+$ filament within the disk cavity around HD 142527.  
\citet{casassus13} discovered this filament to be oriented nearly radially; in 
estimating $\dot{M}$ for this system, they assumed $v_R \sim v_{\rm K}$ 
(although the velocities of the HCO$^+$ filament were not actually quoted in
their study).

>>>>

\chapter{Fast Radial Flows in Transition Disk Holes}\label{chap:radial_flows}

%%\section*{abstract}

@@

A subset of T Tauri and Herbig Ae disks have large, mass-depleted cavities in 
their inner regions, inferred from their infrared spectra 
\citep[e.g.,][]{Kim13} or spatially resolved imaging at infrared 
\citep[e.g.,][]{geers07} and mm wavelengths \citep[e.g.,][]{brown09}.  The
cavities in these ``transition'' disks have radii ranging from $\sim$4\,AU 
\citep[TW Hya;][]{calvet02,hughes07} to 20\,AU \citep[GM Aur;][]{hughes09} to 
40--50\,AU \citep[LkCa 15;][]{andrews11b,isella12} to 140\,AU \citep[HD 
142527;][]{casassus13}.  The suspicion is that dynamical interactions with 
nascent planetary systems evacuate the inner disk regions 
\citep[e.g.,][]{Zhu11}.

@@

Disk radial velocities driven by embedded massive planets or brown dwarfs can 
be healthy fractions of orbital (azimuthal) velocities.  We derive a crude 
estimate\footnote{Our ``zero-dimensional'' analysis ignores the fact that 
actual disk flows around planets are two-dimensional and dominated by
non-axisymmetric streams flowing in and around planetary Hill spheres.
Still, some indirect empirical support for the scalings in equation 
(\ref{eqn:simple}) has been found by \cite{fung14}.
}
 for $v_R$ by equating the one-sided Lindblad torque --- the amount of 
angular momentum transmitted outward across a planet's orbit 
\citep{goldreich80} --- with the angular momentum advected inward by accretion, 
$\dot{M} v_{\rm K} R$.  This balance yields
\begin{equation}
v_R \sim \left( \frac{M_p}{M_\ast} \right)^2 \left( \frac{H}{R} \right)^3 v_{\rm K} \label{eqn:simple}
\end{equation}
where we have used the total linear Lindblad torque, integrated over all 
resonances outside the torque cut-off zone at distances $\gtrsim H$ away from 
the planet (see, e.g., \citealt{goodman01} or \citealt{crida06}, and references 
therein).  For a planet mass $M_p \sim 10\,M_{\rm Jup}$, a host stellar mass 
$M_\ast \sim 1\,M_\odot$, and a typical disk aspect ratio $H/R \sim 0.1$, 
equation (\ref{eqn:simple}) evaluates to
\begin{equation}
v_R \sim 0.1 \,v_{\rm K} \sim 0.5 \left( \frac{30\,{\rm AU}}{R} \right)^{1/2}\,{\rm km} \,\, {\rm s}^{-1}. 
\label{eqn:simple1}
\end{equation}
Modern heterodyne receivers on (sub)millimeter wavelength interferometers 
routinely observe disks at $\sim$0.1 km s$^{-1}$ resolution, and the Atacama 
Large Millimeter Array (ALMA) is expected to provide high sensitivity data at 
finer velocity scales yet.  Evidence for free-fall radial velocities can be 
found in the HCO$^+$ filament within the disk cavity around HD 142527.  
\citet{casassus13} discovered this filament to be oriented nearly radially; in 
estimating $\dot{M}$ for this system, they assumed $v_R \sim v_{\rm K}$ 
(although the velocities of the HCO$^+$ filament were not actually quoted in
their study).
